Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-11-28 threshold/draft.Rnw
More draft notes

Including more terms from Equation \ref{vdist} increases our ability to prune
adjacent pairs of vectors below the correlation threshold with respect to the
ordering defined by $P$.  Including only one more term finds that 49\% of
adjacent vectors fall below the a correlation threshold of $0.95$ in the
example shown in Figure \ref{fig1} and are pruned by identifying indices such
that
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 > 2(1-t).
\]
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout. Including ten terms prunes over 96\% of adjacent vector pairs,
leaving only 15 candidate pairs of adjacent vectors that might possibly
meet the correlation threshold.
In general, using $p\le k$ terms to prune pairs of vectors below the
correlation threshold boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 > 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values.

@@

We evaluted the algorithm using three publically-available gene expression
datasets.  The tests were performed on a desktop PC with a single AMD A10-7850K
3.7$\,$GHz Athlon quad-core CPU and 16$\,$GB 1333$\,$MHz unbuffered RAM running
the Ubuntu 12.04.5 GNU/Linux OS. Tests were performed with 64-bit R version
3.1.2 using double-precision arithmetic and the OpenBLAS library version
0.1alpha2.2-3 (based on Goto's BLAS version 1.13) with {\tt
OMP\_NUM\_THREADS=4}.  All examples use the reference R implementation
function, {\tt tcor()}, which can be installed for the R language from the
development version of IRLBA GitHub repository using the {\tt devtools} package
with:

<<eval=FALSE,echo=TRUE>>=
devtools::install_github("bwlewis/irlba")
@

\subsection{BicatYeast small example}

The first example is a tiny proof of concept that provides the data for Figure
\ref{fig1} and illustrates our timing and memory-use measurement methodology.
Anyone can quickly run the expample to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
BicatYeast} data from the {\tt biclust} R package and related {\tt BicAT}
software\cite{biclust, bicat}. The data consist of 70 experiments invovling the
Saccharomyces Cerevisiae yeast measuring gene expression data across 419
Affymetrix probes, represented as a $70\times 419$ real-valued matrix.  The
example finds all pairs of columns (Affymetrix probes) that meet or exceed a
Pearson correlation value of 0.95.  We use a projected dimension $p=10$ in {\tt tcor()}.
\begin{small}
<<eval=FALSE,echo=TRUE>>=
library(irlba)
library(biclust)  # for the BicatYeast data
data(BicatYeast)  # in gene by sample orientation
A = t(BicatYeast) # to correlate genes to genes
threshold = 0.95

@@

cat("tcor time (s)                      ", t1, "\n")
cat("tcor peak excess memory (MB)       ", m1, "\n")

# Compute using R's native cor() function:
m2  = sum(gc()[,2])
t2  = proc.time()
cx  = cor(A)                # full correlation matrix
diag(cx) = -Inf             # ignore diagonal
pairs = which(cx >= threshold, arr.ind=TRUE)
t2  = (proc.time() - t2)[3]
m2  = sum(gc()[,6]) - m2

cat("brute force time (s)               ", t2, "\n")
cat("brute force peak excess memory (MB)", m2, "\n")

# Compare results!
cat("brute force vector pairs found", nrow(pairs)/2, "\n")
cat("tcor vector pairs found       ", nrow(tx[[1]]), "\n")
@
\end{small}
Brute-force computation outperforms {\tt tcor()} in this tiny first example,
both in speed and slightly in memory use.  Each algorithm identifies the same
115 vector pairs that exceed the specified correlation threshold of 0.95, with
comparative timing and memory use shown in the following table.

\begin{center}
\begin{tabular}{l|rr}
 & {\tt tcor()} & native {\tt cor()} \\
Wall-clock time (s) & 0.26 & 0.02 \\
Excess peak memory use (MB) & 29.2 & 26.2 \\
\end{tabular}
\end{center}

@@

Brute-force computation benefits substantially from sustained use of BLAS Level
3 operations (matrix multiplication), achieving high utilization of available
CPU floating point capacity. The tcor method also benefits from optimized BLAS
routines, but not as much as it mostly makes use of Level 2 operations and
Level 3 operations on sequences of smaller problems.

%<<echo=FALSE, results=tex>>=
%require(xtable)
%print(
%    xtable(head(iris),
%        caption = "blah", label = "tab:five", table.placement = "!h", caption.placement = "top")
%  , size = "small", table.placement="ht")
%@

\section{Extensions}\label{extensions}
\section{Conclusion}\label{conclusion}

@@

%% irlba
\bibitem{irlba}
Baglama, James, and Lothar Reichel. Augmented implicitly restarted Lanczos bidiagonalization methods. SIAM Journal on Scientific Computing 27.1 (2005): 19-42.
\bibitem{irlbam}
Baglama, James,  {\url{http://www.math.uri.edu/~jbaglama/software/irlba.m}}.

@@

>>>>

Including more terms from Equation \ref{vdist} increases our ability to prune
adjacent pairs of vectors below the correlation threshold with respect to the
ordering defined by $P$.  Including only one more term finds that 49\% of
adjacent vectors fall below the correlation threshold of $0.95$ in the
example shown in Figure \ref{fig1} and are pruned by identifying indices such
that
\[
\sigma_1^2 (D^1 P v_1)^2 +
\sigma_2^2 (D^1 P v_2)^2 > 2(1-t).
\]
Note that we use the permutation $P$ defined by the order of vector $v_1$
throughout. Including ten terms prunes over 96\% of adjacent vector pairs,
leaving only 15 candidate pairs of adjacent vectors that might possibly
meet the correlation threshold.
In general, using $p\le k$ terms to prune pairs of vectors below the
correlation threshold boils down to evaluating the expression
\begin{equation}\label{proj}
(D^1 P V_{1:p} )^2 s_{1:p}^2 > 2(1-t),
\end{equation}
where 
$V_{1:p}$ denotes the first $p$ columns of the
matrix $V$, $s_{1:p}$ the vector of the first $p$ singular values.

@@

We evaluted the algorithm using three publically-available gene expression
datasets.  The tests were performed on a desktop PC with a single AMD A10-7850K
3.7$\,$GHz Athlon quad-core CPU and 16$\,$GB 1333$\,$MHz unbuffered RAM running
the Ubuntu 12.04.5 GNU/Linux OS. Tests were performed with 64-bit R version
3.1.2 using double-precision arithmetic and the OpenBLAS library version
0.1alpha2.2-3 (based on Goto's BLAS version 1.13) with {\tt
OMP\_NUM\_THREADS=4}.  All examples use the reference R implementation
function, {\tt tcor()}, which can be installed for the R language from the
development GitHub repository using the {\tt devtools} package with:

<<eval=FALSE,echo=TRUE>>=
devtools::install_github("bwlewis/tcor")
@

The native R {\tt cor()} function is a general function that can compute
several different types of correlation matrices including Pearson, Spearman,
and Kendall and includes a number of options for dealing with missing values.
Because of its flexibility, the native {\tt cor()} function does not always
compute Pearson correlation matrices in the most computationally-efficient way.

Correlation matrix computation can benefit substantially from use of optimized
BLAS Level 3 operations (matrix multiplication), achieving high utilization of
available CPU floating point capacity. The {\tt tcor()} function also benefits
from optimized BLAS routines, but not as much as it mostly makes use of Level 2
operations and Level 3 operations on sequences of smaller problems.

We compare the {\tt tcor()} function with brute-force Pearson correlation
matrix computation computed in the fastest way we can, directly using matrix
multiplication instead of using the slower {\tt cor()} function.

\subsection{Small BicatYeast example}

The first example is a tiny proof of concept that provides the data for Figure
\ref{fig1} and establishes our timing and memory use measurement methodology.
Wall-clock times are reported in seconds and peak memory use in excess of
memory required to store the input data matrix is reported in megabytes.

Anyone can quickly run the expample to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
BicatYeast} data from the {\tt biclust} R package and related {\tt BicAT}
software\cite{biclust, bicat}. The data consist of 70 experiments invovling the
Saccharomyces Cerevisiae yeast measuring gene expression data across 419
Affymetrix probes, represented as a $70\times 419$ real-valued matrix.  The
example finds all pairs of columns (Affymetrix probes) that meet or exceed a
Pearson correlation value of 0.95.  We use a projected dimension $p=10$ in {\tt tcor()}.
\begin{small}
<<eval=TRUE,echo=TRUE>>=
library(irlba)
source("tcor.R")  # XXX Replace later with tcor pacakge
library(biclust)  # for the BicatYeast data
data(BicatYeast)  # in gene by sample orientation
A = t(BicatYeast) # to correlate genes to genes
threshold = 0.95

@@

# Fast brute force full correlation:
m2  = sum(gc()[,2])
t2  = proc.time()
mu = colMeans(A)
s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
cs = scale(A, center=mu, scale=s)  # scale and center the data
cx = crossprod(cs)                 # full correlation matrix
cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
pairs = which(cx >= threshold, arr.ind=TRUE)
t2  = (proc.time() - t2)[3]
m2  = sum(gc()[,6]) - m2
@
\end{small}
Brute-force computation outperforms {\tt tcor()} in this tiny first example,
both in speed and slightly in memory use.  Each algorithm identifies the same
115 unique vector pairs that exceed the specified correlation threshold of
0.95, with comparative timing and memory use shown in the following table.

<<echo=FALSE, results=tex>>=
require(xtable)
ans = data.frame(Method=c("tcor","cor"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "BicatYeast results, threshold=0.95", label = "BicatYeast", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

\subsection{TCGA Gene Expression Examples}

Performance benefits from the {\tt tcor()} function become apparent whith large
data.  The next example uses RPKM-normalized gene expression data measured
across 20,532 genes from 878 breast cancer tumor samples obtained from the
Broad GDAC dashboard\cite{gdac} of the Cancer Genome Atlas data. See the
supplemental code at XXX  (also embedded in this Sweave document) for examples
of efficiently converting the downloaded sample expression data into an R
matrix.

Eight columns with constant zero values were removed from the data in a
pre-processing step, leaving an $878\times 20524$ matrix. The example computes
pairs of gene expression column vectors with correlation values of 0.99
or more. We use the same {\tt tcor} projection dimension of 10 and performance
measurement methodology as used in the first example.

Both methods find the same 3,572 pairs of vectors. The fast brute-force
computation is slightly faster than the {\tt tcor()} function largely due to
the still relatively small input matrix and better utilization of optimized
BLAS routines. But the {\tt tcor()} function uses far less memory.  (For
comparison, R's {\tt cor()} function took about 372 seconds and about
3200$\,$MB RAM on the same problem on the same test system.)

<<echo=FALSE, results=tex>>=
# --------------------------------------------------------------------------------------------------
# The first time you run this, you need to download the data as outlined in the
# following commented-out section. This code creates a file named
# 'rpkm.RData' used subsequently.
#
# To run any of the remaining (large) examples, first set the system environment
# variable RUN_EXAMPLES=TRUE (otherwise they won't run and representative results
# from Bryan's test machine will be printed).
#
#library(doMC)
# Download and extract this file with, for example:
# wget  http://gdac.broadinstitute.org/runs/stddata__2015_08_21/data/BRCA/20150821/gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0.tar.gz
# tar zxf gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0.tar.gz
# cd gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0
# R --slave < read.R   # (run this file)
#f = "BRCA.rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.data.txt"
#N = 20534  # total number of lines
#h = 2      # header lines
# Read the first header line with sampe IDs:
#id = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=1))
# Read the 2nd header line with sample type (raw, RPKM, etc.)
#header = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, skip=1, header=FALSE, nrows=1))
# Read the rest of the file in parallel. I prefer doMC/foreach over mclapply
# for this because foreach can combine results incrementally.
#cores = 4
#registerDoMC(cores)
#block = floor((N - h)/cores)
#t1 = proc.time()
#ans = foreach(j=1:cores, .combine=rbind) %dopar%
#{
#    skip = block * (j - 1) + h
#    nrows = ifelse(j == cores, -1, block)
#    read.table(f, sep="\t", stringsAsFactors=FALSE, skip=skip, header=FALSE, nrows=nrows)
#}
#names(ans) = header
#print(proc.time() - t1)
# (All timings shown for a quad-core Athlon home PC reading from an older
# OCZ REVODRIVE3 X2 SSD drive (ca 2014).)
#   user  system elapsed
#494.549   4.134 131.232
# Compare with sequential approach:
#t1 = proc.time()
#x = read.table(f, sep="\t", stringsAsFactors=FALSE, skip=1, header=TRUE)
#print(proc.time() - t1)
#   user  system elapsed
#517.430   1.318 519.553
# Create a samples x gene matrix of RPKM values
#genes = ans[, 1] # gene labels
#idx   = grep("RPKM", names(ans))    # Select columns with RPKM values
#rpkm     = t(as.matrix(ans[, idx]))
# Label the coordinates by sample ID and gene:
#rownames(rpkm) = id[idx]
#colnames(rpkm) = genes
#save(rpkm, file="rpkm.RData")
# --------------------------------------------------------------------------------------------------
# Default values:
t1=40.34
t2=35.60
m1=39.10
m2=9595.10

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
load("rpkm.RData")  # rpkm expression matrix

# The data have a few columns of all zeros that need to be removed
n = ncol(rpkm)
rpkm = rpkm[, colMeans(rpkm) > 0]

threshold = 0.99     # correlation threshold
xx  = gc()           # clear
m1  = sum(gc()[,2])  # Memory in megabytes currently in use
t1  = proc.time()
tx  = tcor(rpkm, threshold, 11)
t1  = (proc.time() - t1)[3] # Time in seconds
m1  = sum(gc()[,6] - m1)    # Peak excess memory use during the test

# Compare with the fastest way we know of to compute a thresholded correlation
# matrix by brute force. This approach makes heavy use of fast multi-threaded
# BLAS routines. This approach is much faster than R's default `cor` function.
xx  = gc()
m2  = sum(gc()[,2])  # Memory in megabytes currently in use
t2 = proc.time()
mu = colMeans(rpkm)
s  = sqrt(apply(rpkm, 2, crossprod) - nrow(rpkm) * mu ^ 2) # col norms of centered matrix
cs = scale(rpkm, center=mu, scale=s)  # scale and center the data
bx = crossprod(cs)                    # full correlation matrix
diag(bx) = 0                          # ignore diagonal (all ones)
pairs = which(bx >= threshold, arr.ind=TRUE)
t2  = (proc.time() - t2)[3] # Time in seconds
m2  = sum(gc()[,6] - m2)    # Peak excess memory use during the test
} # RUN_EXAMPLES=TRUE

require(xtable)
ans = data.frame(Method=c("tcor","cor"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA BRCA gene expression results (878 samples x 20,542 genes, threshold=0.99)", label = "TCGA1", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

We simulated a data set with a larger number of samples by simply replicating
the rows of the gene expression matrix 27 times to obtain a matrix with 23,706
rows and 20,524 columns, noting that the simulated data have the same rank (of
at most 878) and the same expected number of vector pairs meeting the
correlation threshold as the original data. Each method again computed the same
set of  3,572 pairs of vectors metting the correlation threshold of 0.99, but
the {\tt tcor()} function outperforms the brute force method in both compute
time and memory consumption.

<<echo=FALSE, results=tex>>=
# Default values:
t1=372.69
t2=546.06
m1=768.3
m2=9600.2

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
  # simulated data, replicate rows 27x
}
require(xtable)
ans = data.frame(Method=c("tcor","cor"), `Wall-clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA simulated gene expression results (23,706 rows x 20,542 columns, threshold=0.99)", label = "TCGA2", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

@@

\section{Extensions}\label{extensions}

XXX Two basic extensions, include functions/options for this in the package.

The main ideas in Algorithm \ref{prune}, although applied to correlation,
easily extend to any distance metric.

Algorithm \ref{prune} can be adapted to find pairs of highly anti-correlated
vectors by changing the distance metric...XXX

\section{Conclusion}\label{conclusion}

@@

%% irlba
\bibitem{irlba}
Baglama, James, and Lothar Reichel. Augmented implicitly restarted Lanczos bidiagonalization methods. SIAM Journal on Scientific Computing 27.1 (2005): 19-42.

\bibitem{irlbam}
Baglama, James,  {\url{http://www.math.uri.edu/~jbaglama/software/irlba.m}}.

@@

%% BROAD GDAC dashboard
\bibitem{gdac}{\url{http://gdac.broadinstitute.org/runs/stddata__2015_08_21/data/BRCA/20150821/gdac.broadinstitute.org_BRCA.Merge_rnaseq__illuminahiseq_rnaseq__unc_edu__Level_3__gene_expression__data.Level_3.2015082100.0.0.tar.gz}}
